### 2015-09-28 chapters/orthography_profiles.tex
update formal specifications

The formal specifications of an orthography profile (or simply \textsc{profile}
for short) are the following:

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section~\ref{pitfall-file-formats}, Pitfall: File Formats) that includes the information pertinent to the orthography. 
	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}. A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header. For all columns we assume the name in the header of the CSV file to be crucial. The actual ordering of the columns is unimportant. 
	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash \textsc{\#} (\textsc{number sign} at \uni{0023}). Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}. Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.} 
	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic \textsc{tag: value} format. }Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile. 
\end{enumerate}

The content of a profile consists of lines, each describing a grapheme of the
orthography, using the following columns:

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{A minimal profile consists of a single column with header \textsc{graphemes}}, listing each of the different graphemes in a separate line. 
	\item \textsc{Optional columns called \textsc{left} and \textsc{right} can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \textsc{grapheme}, \textsc{left} and \textsc{right} can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops <.> or dollar signs <\$> (so-called \textsc{metacharacters}) have to be explicitly escaped by adding a backslash before them (i.e.~use <.> or <\$>). Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * . - ! ? \^{} \$ . 
	\item \textsc{An optional column called \textsc{class} can be used to specify classes of graphemes}, for example to define a class of vowels. Users can simply add ad-hoc identifiers in this column to indicate a group of graphemes, which can then be used in the description of the graphemes or the context. The identifiers should of course be chosen such that they do not conflate with any symbols used in the orthography themselves. Note that such classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be added and named at will}. Often more than a single possible transliteration will be of interest. Any software application using these profiles should use the names of these columns to select a specific transliteration column. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored by any software application using the profiles}. As orthography profiles are also intended to be read and interpreted by humans, it is often highly useful to add extra information on the graphemes in further columns, like for example Unicode codepoints, Unicode names, frequency of occurrence, examples of occurrence, explanation of the contextual restrictions, or comments. 
\end{enumerate}

For the automatic processing of the profiles, the following technical standards
will be expected:

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular expression. }Software applications using profiles can also offer to interpret a profile in the literal sense to avoid the necessity for the user to escape regular expressions metacharacters in the profile. However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit \textsc{or} chains of regular expressions}, which will then be inserted in the \textsc{graphemes}, \textsc{left} and \textsc{right} columns at the position indicated by the class-identifiers. For example, a class \textsc{V} as a context specification might be replaced by a regular expression like: (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au). Only the graphemes themselves are included here, not any contexts specified for the elements of the class. 
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included into the regular expressions by using lookbehind and lookahead}. Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves. Internally, the contexts in the columns \textsc{left} and \textsc{right} are combined with the column \textsc{graphemes} to form a complex regular expression like: (?\textless{}=left)graphemes(?=right). 
	\item \textsc{The regular expressions will be applied in the order as specified in the profile, from top to bottom.} A software implementation can offer help in figuring out the optimal ordering of the regular expressions, but should then explicitly report on the order used. 
\end{enumerate}

The actual implementation of the profile on some text-string will function as
follows:

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{All graphemes are matched in the text before they are tokenized or transliterated}. In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (feeding) or prevents another rule to apply (`bleeding'). 
	\item \textsc{The matching of the graphemes can occur either globally or linearly. }From a computer science perspective, the most natural way to match graphemes from a profile in some text is by walking linearly through the text-string from left to right, and at each position go through all graphemes in the profile to see which one matches, then go to the position at the end of the matched grapheme and start over. This is basically how a finite state transducer works, which is a well-established technique in computer science. However, from a linguistic point of view, our experience is that most linguists find it more natural to think from a global perspective. In this approach, the first grapheme in the profile is matched everywhere in the text-string first, before moving to the next grapheme in the profile. Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result. Still, we suggest that any software application using orthography profiles should offer both approaches (i.e. \textsc{global} or \textsc{linear}) to the user. The approach used should be documented in the metadata as \textsc{tokenization method}. 
	\item \textsc{The matching of the graphemes can occur either in NFC or NFD. }By default, both the profile and the text-string to be tokenized should be treated as NFC (see section \ref{pitfall-canonical-equivalence}, Pitfall: Canonical equivalence, above). However, in some use-cases it turns out to be practical to treat both text and profile as NFD. This typically happens when very many different combinations of diacritics occur in the data. An NFD-profile can then be used to first check which individual diacritics are used, before turning to the more cumbersome inspection of all combinations. We suggest that any software application using orthography profiles should offer both approaches (i.e. \textsc{NFC} or \textsc{NFD}) to the user. The approach used should be documented in the metadata as \textsc{unicode normalization}. 
	\item \textsc{The text-string is always returned in tokenized form} by separating the matched graphemes by a user-specified symbols-string. Any transliteration will be returned on top of the tokenization. 
	\item \textsc{Leftover characters (i.e.~characters that are not matched by the profile) should be reported to the user as errors.} Typically, the unmatched character are replaced in the tokenization by a user-specified symbol-string. 
\end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally, a warning might be given if the chosen string includes characters from the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed \textsc{global} or \textsc{linear}. 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use \textsc{NFC} or \textsc{NFD}. 
	\end{enumerate}
	\item \textsc{might offer user-options }to:
	\begin{enumerate}
		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{assist in the ordering of the graphemes.} In our experience, it makes sense to apply larger graphemes before shorter graphemes, and to apply graphemes with context before graphemes without context. Further, frequently relevant rules might be applied after rarely relevant rules (though frequency is difficult to establish in practice, as it depends on the available data). Also, if this all fails to give any decisive ordering between rules, it seems useful to offer linguists the option to reverse the ordering from any manual specified ordering, because linguists tend to write the more general rule first, before turning to exceptions or special cases. 
		\item \textsc{assist in dealing with upper and lower case characters.} It seems practical to offer some basic case matching, so characters like <a> and <A> are treated equally. This will be useful in many concrete cases, although the user should be warned that case matching does not function universally in the same way across orthographies. Ideally, users should prepare orthography profiles with all lowercase and uppercase variants explicitly mentioned, so by default no case matching should be performed. 
		\item \textsc{treat the profile literal}, i.e.~to not interpret regular expression metacharacters. Matching graphemes literally often leads to strong speed increase, and would allow users to not needing to worry about escaping metacharacters. However, in our experience all actually interesting use-cases of orthography profiles include some contexts, which automatically prevents any literal interpretation, so by default the matching should not be literal. 
	\end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings to be processed in the used Unicode normalization}, i.e.~in either NFC or NFD as specified by the user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated
        strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both in which
        text-strings any errors occurred and which characters in the
        text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered 
	\end{enumerate}
\end{enumerate}

>>>>

\subsection*{File Format}
The formal specifications of an orthography profile (or simply \textsc{profile}
for short) are the following:

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} 
	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
       includes the information pertinent to the orthography. 
	\item \textsc{A profile is a tab-separated CSV file with an obligatory header
       line}. A minimal profile can have just a single column, in which case
       there will of course be no tabs, but the first line will still be the
       header. For all columns we assume the name in the header of the CSV file
       to be crucial. The actual ordering of the columns is unimportant. 
	\item \textsc{Metadata are added in a separate UTF-8 text file} using a basic
       \textsc{tag: value} format. Metadata about the orthographic description
       given in the orthography profile includes, minimally, (i) author, (ii)
       date, (iii) title of the profile, (iv) a stable language identifier
       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
       bibliographic data for resource(s) that illustrate the orthography
       described in the profile. Further, (vi) the tokenization method and (vii)
       the unicode normalisation used should be documented here (see below).
    	\item \textsc{Separate lines with comments are not allowed}. Comments that
       belong to specific lines will have to be put in a separate column of
       the CSV file, e.g.~add a column called \textsc{comments}.            
\end{enumerate}

\noindent The content of a profile consists of lines, each describing a grapheme
of the orthography, using the following columns:

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right}, respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression metacharacters.} If regular expressions are
       used, then all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is probably
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen such that they do not conflate
       with any symbols used in the orthography themselves. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       highly useful to add extra information on the graphemes in further
       columns, like for example Unicode codepoints, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of the contextual
       restrictions, or comments. 
 \end{enumerate}

\subsection*{Implementation}
For the automatic processing of the profiles, the following technical standards
will be expected:

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression. }Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expressions metacharacters in the profile.
       However, this only is possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       should then explicitly report on the order used.      
\end{enumerate}

\noindent The actual implementation of the profile on some text-string will function as
follows:

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{All graphemes are matched in the text before they are tokenized
       or transliterated}. In this way, there is no necessity for the user to
       consider `feeding' and `bleeding' situations, in which the application of
       a rule either changes the text so another rule suddenly applies (feeding)
       or prevents another rule to apply (`bleeding'). 
	\item \textsc{The matching of the graphemes can occur either globally or
       linearly. }From a computer science perspective, the most natural way to
       match graphemes from a profile in some text is by walking linearly
       through the text-string from left to right, and at each position go
       through all graphemes in the profile to see which one matches, then go to
       the position at the end of the matched grapheme and start over. This is
       basically how a finite state transducer works, which is a
       well-established technique in computer science. However, from a
       linguistic point of view, our experience is that most linguists find it
       more natural to think from a global perspective. In this approach, the
       first grapheme in the profile is matched everywhere in the text-string
       first, before moving to the next grapheme in the profile. Theoretically,
       these approaches will lead to different results, though in practice of
       actual natural language orthographies they almost always lead to the same
       result. Still, we suggest that any software application using orthography
       profiles should offer both approaches (i.e. \textsc{global} or
       \textsc{linear}) to the user. The approach used should be documented in
       the metadata as \textsc{tokenization method}. 
	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.} By
       default, both the profile and the text-string to be tokenized should be
       treated as NFC (see Section~\ref{pitfall-canonical-equivalence}). However, in some use-cases it turns out to
       be practical to treat both text and profile as NFD. This typically
       happens when very many different combinations of diacritics occur in the
       data. An NFD-profile can then be used to first check which individual
       diacritics are used, before turning to the more cumbersome inspection of
       all combinations. We suggest that any software application using
       orthography profiles should offer both approaches (i.e. \textsc{NFC} or
       \textsc{NFD}) to the user. The approach used should be documented in the
       metadata as \textsc{unicode normalization}. 
	\item \textsc{The text-string is always returned in tokenized form} by
       separating the matched graphemes by a user-specified symbols-string. Any
       transliteration will be returned on top of the tokenization. 
	\item \textsc{Leftover characters}, i.e.~\textsc{characters that are not
       matched by the profile, should be reported to the user as errors.}
       Typically, the unmatched character are replaced in the tokenization by a
       user-specified symbol-string.       
 \end{enumerate}

\subsection*{Software applications}

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
        a warning might be given if the chosen string includes characters from
        the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
        tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
        proceed \textsc{global} or \textsc{linear}. 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
        should use \textsc{NFC} or \textsc{NFD}. 
    \end{enumerate}
	\item \textsc{might offer user-options }to:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{assist in the ordering of the graphemes.} In our experience, it
        makes sense to apply larger graphemes before shorter graphemes, and to
        apply graphemes with context before graphemes without context. Further,
        frequently relevant rules might be applied after rarely relevant rules
        (though frequency is difficult to establish in practice, as it depends
        on the available data). Also, if this all fails to give any decisive
        ordering between rules, it seems useful to offer linguists the option to
        reverse the ordering from any manual specified ordering, because
        linguists tend to write the more general rule first, before turning to
        exceptions or special cases. 
		\item \textsc{assist in dealing with upper and lower case characters.} It
        seems practical to offer some basic case matching, so characters like
        <a> and <A> are treated equally. This will be useful in many concrete
        cases, although the user should be warned that case matching does not
        function universally in the same way across orthographies. Ideally,
        users should prepare orthography profiles with all lowercase and
        uppercase variants explicitly mentioned, so by default no case matching
        should be performed. 
		\item \textsc{treat the profile literal}, i.e.~to not interpret regular
        expression metacharacters. Matching graphemes literally often leads to
        strong speed increase, and would allow users to not needing to worry
        about escaping metacharacters. However, in our experience all actually
        interesting use-cases of orthography profiles include some contexts,
        which automatically prevents any literal interpretation.
    \end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings to be processed in the specified
        unicode normalization}, i.e.~in either NFC or NFD as specified by the
        user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated
        strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both in which
        text-strings any errors occurred and which characters in the
        text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered 
	\end{enumerate}
\end{enumerate}
